---
title: ""
---

<br><br>

### Setup

Please ensure that you have the latest version of [R](https://www.r-project.org) installed on your laptop. This is important because some R packages used during the course may not install correctly if R is not up to date.     

*macOS users*   
Install R by downloading and running this [.pkg](http://cran.r-project.org/bin/macosx/R-latest.pkg) file from [CRAN](https://cran.r-project.org/index.html).       
 
*Windows users*    
Install R by downloading and running this [.exe](https://cran.r-project.org/bin/windows/base/release.htm) file from [CRAN](https://cran.r-project.org/index.html).      

Please also install [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/). RStudio is a free and open-source integrated development environment (IDE) for R.

<br>

**Testing whether the installation was successful**

Copy and paste the R code below into [RStudio's console pane](http://r4ds.had.co.nz/screenshots/rstudio-layout.png). Then hit Enter to run it. 
```{r, eval=FALSE}
install.packages("tidyverse")
library(ggplot2)
ggplot(mtcars, aes(mpg, wt)) +
  geom_point()
```

If the plot below appears in the output pane of RStudio the installation was successful.   
<br>

```{r, echo=FALSE, fig.height=3}
library(ggplot2)
ggplot(mtcars, aes(mpg, wt)) +
  geom_point()
```

<br>