---
title: | 
    | Getting Started in R
    | an introduction to data analysis and visualisation
subtitle: "Import"
author: "Réka Solymosi & Sam Langton"
date: "2 July 2019"
output:
  xaringan::moon_reader:
    css: ["default", "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse) 
```

class: inverse, center, middle

# Import

---
### Sources of data

- Files stored on your computer
- Web link
- Database
- Web page
- Application Programming Interface (API)

---
### Data formats

.pull-left[
#### Rectangular
- CSV (.csv)
- Excel files (.xls, .xlsx)
- SPSS, Stata, and SAS files (.sav, .dta, .sas)
]

.pull-right[
#### Hierarchical
- eXtensible Markup Language (.xml)
- JavaScript Object Notation (.json)
- Geographic JavaScript Object Notation (.geojson)
]

---
### Reading in data

.pull-left[
#### Rectangular
#### .csv

```r
library(readr)
read_csv()
```

#### .xls, .xlsx

```r
library(readxl)
read_excel()
```

#### .sav, .dta, .sas

```r
library(haven)
read_sav()
read_dta()
read_sas()
```
]

.pull-right[
#### Hierarchical
#### .xml
```r
library(xml2)
read_xml()
```

#### .json
```r
library(jsonlite)
read_json()
```

#### .geojson
```r
library(sf)
st_read()
```
]

---
class: inverse, middle, center

# Parse

---
### Parsing data
The process of ...

1. converting a text file into a matrix of rows and columns and,
2. coercing columns to different data type(s)

---
#### .csv, .xls, .xlsx

```{r comment = '#', tidy = FALSE}
library(readr)
read_csv(
"country,continent,year,lifeExp,pop,gdpPercap\
Italy,Europe,2007,80.546,58147733,28569.7197\
Nigeria,Africa,2007,46.859,135031164,2013.977305"
)
```

---
#### .xml
.pull-left[
```{r comment = '#', tidy = FALSE}
library(xml2)
xml <- 
'<?xml version="1.0" encoding="UTF-8"?>
<root>
  <row>
    <country>Iceland</country>
    <continent>Europe</continent>
    <year>2007</year>
    <lifeExp>81.757</lifeExp>
    <pop>301931</pop>
    <gdpPercap>36180.78919</gdpPercap>
  </row>
  <row>
    <country>Italy</country>
    <continent>Europe</continent>
    <year>2007</year>
    <lifeExp>80.546</lifeExp>
    <pop>58147733</pop>
    <gdpPercap>28569.7197</gdpPercap>
  </row>
</root>'
```
]
.pull-right[
```{r comment = '#', tidy = FALSE}
xml_obj <- read_xml(xml) #<<
country <- xml_text(xml_find_all(xml_obj, "//country")) #<<
continent <- xml_text(xml_find_all(xml_obj, "//continent")) #<<
year <- xml_text(xml_find_all(xml_obj, "//year")) #<<
lifeExp <- xml_text(xml_find_all(xml_obj, "//lifeExp")) #<<
pop <- xml_text(xml_find_all(xml_obj, "//pop")) #<<
gdpPercap <- xml_text(xml_find_all(xml_obj, "//gdpPercap")) #<<
data.frame(country, continent, year, lifeExp, pop, gdpPercap) #<<
```
]

---
#### .json

```{r comment = '#', tidy = FALSE, message = FALSE}
library(jsonlite)
json <- 
'[
  {
    "country": "Italy",
    "continent": "Europe",
    "year": 2007,
    "lifeExp": 80.546,
    "pop": 58147733,
    "gdpPercap": 28569.7197
  },
  {
    "country": "Nigeria",
    "continent": "Africa",
    "year": 2007,
    "lifeExp": 46.859,
    "pop": 135031164,
    "gdpPercap": 2013.977305
  }
]'
fromJSON(json) #<<
```

---
#### .geojson
.pull-left[
```{r comment = '#', tidy = FALSE, message = FALSE, fig.height = 4, dev = 'png'}
library(sf)
italy <- st_read("https://code.highcharts.com/mapdata/countries/it/it-all.geo.json", quiet = TRUE) #<<
as.data.frame(italy)[1:5, 14:15]
plot(st_geometry(italy))
```
]

.pull-right[
```{r comment = '#', tidy = FALSE, message = FALSE, fig.height = 4, dev = 'png'}
nigeria <- st_read("https://code.highcharts.com/mapdata/countries/ng/ng-all.geo.json", quiet = TRUE) #<<
as.data.frame(nigeria)[1:5, 14:15]
plot(st_geometry(nigeria))
```
]

<!-- --- -->
<!-- ### Coercing columns to different data type -->

<!-- - `logical` variables can store TRUE or FALSE -->
<!-- - `integer` variables can only store whole numbers -->
<!-- - `double` variables can store decimal numbers -->
<!-- - `character` variables can store strings of characters -->
<!-- - `date` variables can store a date - time -->
<!-- - `factor` variables can store a fixed and known set of possible values -->

---
#### .csv, .xls, .xlsx

```{r comment = '#', tidy = FALSE}
library(readr)
df <- read_csv(
"country,continent,year,lifeExp,pop,gdpPercap\
Italy,Europe,2007,80.546,58147733,28569.7197\
Nigeria,Africa,2007,46.859,135031164,2013.977305",
col_types = cols( #<<
  country = col_character(), #<<
  continent = col_character(), #<<
  year = col_integer(), #<<
  lifeExp = col_double(), #<<
  pop = col_integer(), #<<
  gdpPercap = col_double() #<<
  ) #<<
)
glimpse(df)
```

---
# Export

---
### Writing out data

.pull-left[
#### Rectangular
#### .csv

```r
library(readr)
write_csv()
```

#### .xls, .xlsx

```r
library(readxl)
write_excel()
```

#### .sav, .dta, .sas

```r
library(haven)
write_sav()
write_dta()
write_sas()
```
]

.pull-right[
#### Hierarchical
#### .xml
```r
library(xml2)
write_xml()
```

#### .json
```r
library(jsonlite)
write_json()
```

#### .geojson
```r
library(sf)
st_write()
```
]

---
### Please note

- R stores factors as integer codes 1, 2, 3 etc. not character strings. So ensure that you set your factor levels.
- Column type information is lost when you export to a CSV. Recommend exporting to .Rds, the native R data format using `write_rds()`



---
<!-- --- -->
<!-- ### Defining dates and times -->
<!-- - Coerce variables as dates when you need to create a plot, extract information or conduct time series analysis. -->

<!-- .pull-left[ -->
<!-- ```{r comment = '#', tidy = FALSE} -->
<!-- parse_date("03-07-18", "%d-%m-%y") -->
<!-- parse_date("03-07-18", "%m-%d-%y") -->
<!-- parse_date("03-07-18", "%y-%m-%d") -->
<!-- # as.Date(x, "%d-%m-%y") -->
<!-- ``` -->
<!-- ] -->

<!-- .pull-right[ -->
<!-- | Symbol | Description | Example  | -->
<!-- |:--- |:---- |:---- | -->
<!-- |%d| day as a number | 02 | -->
<!-- |%a| abbreviated day | Tue | -->
<!-- |%A| day | Tuesday | -->
<!-- |%m| month as a number | 07 | -->
<!-- |%b| abbreviated month | Jul | -->
<!-- |%B| month | July | -->
<!-- |%y| 2-digit year | 18 | -->
<!-- |%Y| 4-digit year |  2018 | -->
<!-- ] -->

<!-- --- -->
<!-- ### Defining factors -->
<!-- - Coerce variables to factors when you need a specific ordering of string values, e.g. in a model or a plot. -->

<!-- ```{r comment = '#', tidy = FALSE} -->
<!-- parse_factor(df$country, levels = c("Italy", "Nigeria")) -->

<!-- df$country <- factor(df$country,  -->
<!--                      levels = unique(df$country), -->
<!--                      ordered = FALSE) -->
<!-- ``` -->

<!-- --- -->
<!-- class: inverse, middle, center -->

<!-- # Delving into factors -->

<!-- --- -->
<!-- ### Inspecting factors -->
<!-- ```{r comment = '#', tidy = FALSE} -->
<!-- library(gapminder) -->
<!-- class(gapminder$continent) -->
<!-- levels(gapminder$continent) -->
<!-- ``` -->

<!-- --- -->
<!-- ### Dropping unused factor levels -->
<!-- ```{r comment = '#', tidy = FALSE} -->
<!-- g7 <- gapminder %>%  -->
<!--   filter(country %in% c("Canada", "France", "Germany", "Italy", "Japan", "United Kingdom", "United States")) -->
<!-- levels(g7$country)[1:20] -->
<!-- g7$country <- droplevels(g7$country) -->
<!-- levels(g7$country) -->
<!-- ``` -->

<!-- --- -->
<!-- ### Recoding factors -->
<!-- ```{r comment = '#', tidy = FALSE} -->
<!-- library(forcats) -->
<!-- g7$country <- g7$country %>% -->
<!--   fct_recode("UK" = "United Kingdom", "USA" = "United States")  -->
<!-- levels(g7$country) -->
<!-- ``` -->

<!-- --- -->
<!-- ### Reordering factors -->

<!-- ```{r comment = '#', tidy = FALSE, fig.height = 4, dev = 'png'} -->
<!-- g7_2007 <- g7 %>% filter(year == 2007) -->
<!-- levels(g7_2007$country) -->

<!-- ggplot(g7_2007, aes(x = gdpPercap, y = country)) + -->
<!--   geom_point() + -->
<!--   labs(x = "GDP per capita", y = NULL) + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- --- -->
<!-- ### Reordering factors -->
<!-- ```{r comment = '#', tidy = FALSE, fig.height = 4, dev = 'png'} -->
<!-- fct_reorder(g7_2007$country, g7_2007$gdpPercap, .desc = TRUE) %>%  -->
<!--   levels() -->

<!-- ggplot(g7_2007, aes(x = gdpPercap, y = fct_reorder(country, gdpPercap))) + -->
<!--   geom_point() + -->
<!--   labs(x = "GDP per capita", y = NULL) + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- --- -->
<!-- class: inverse, middle, center -->