---
title: | 
    | Getting Started in R
    | an introduction to data analysis and visualisation
subtitle: "Modeling Data"
author: "Réka Solymosi, Sam Langton & Emily Buehler"
date: "4 July 2019"
output:
  xaringan::moon_reader:
    css: ["default", "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: inverse, center, middle

# Modeling Data

---
## Making inferences from data


---
## Univariate 

### Summary of one variable


---
## Univariate 

For all variables in dataframe

```{r, echo=FALSE}
library(gapminder)
```

```{r}
summary(gapminder)
```

---
## Univariate 

For all variables in dataframe with skimr

```{r}
library(skimr)
```

---

```{r}
gapminder %>%  skim()
```


---
## Bivariate stats 

### making comparisons

---
##Numeric - numeric

```{r, eval=FALSE}
cor( var1, var2, method = "method")
```

```{r, eval=FALSE}
cor.test(var1, var2, method = "method")
```

---
##Categorical - categorical

###Cross tabs

```{r, eval=FALSE}

chisq.test(table(var1, var2))

```

```{r, eval=FALSE}
fisher.test(table(var1, var2))
```

---
##Categorical - categorical

###Odds ratio

```{r, eval = FALSE}
library(vcd)

oddsratio(table(var1, var2)) 
```

---
##Categorical - numeric

```{r, eval = FALSE}

#t-test
t.test(var1 ~ var2)

#anova
aov(var1 ~ var2, data=mydataframe)


```

---
##Multivariate 

```{r, eval = FALSE}

fit_1 <- glm(dependent_var ~ independent_var_1 + independent_var_2, data = mydata)
summary(fit_1)

```

---
##Multivariate 

```{r}
fit_2 <- glm(lifeExp ~ continent + gdpPercap, data = gapminder)
summary(fit_2)

```

---
##Multivariate 

```{r}
library(car)

residualPlots(fit_2)
```

---
##Multivariate 

```{r}
marginalModelPlots(fit_2)

```

---
##Multivariate 

```{r, warning=FALSE, message=FALSE}
library(effects)
plot(allEffects(fit_2))
```

---
##Multivariate 

```{r, warning=FALSE, message=FALSE}
library(sjPlot)
plot_model(fit_2, show.values = TRUE)
```



